---
title: "EBR identification"
author: "Filip Thörn"
date: '2025-11-04'
output: html_document
---

#required packages 
```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(data.table)
library(ggtree)
library(GenomicRanges)
library(cowplot)
```

#EBR notes
Make a look up table of from 0 to fasta index end of each chromosome for all buscos counting from 5' to 3' end. 
The breakpoints should go from last the BUSCO of the synteny block (of each scaffold) to the next busco on the reference chromosome. 
The same should be done for the first Busco on the following synteny block but in the opposit direction. 
Keep track of the direction of the synteny block on the reference (count one previous row and one following row).

In the map files the buscos should be numbered 1 to n, from 5' to 3' end.

Each detection of a breakpoint should save the busco on either synteny block and count the next busco on the reference (As described above)

#READ DATA and define subsets
```{r}
#setwd("./")
#Lists of samples used to read all gene lists. Used in 
all_species<-c("LjuvernicaSweF","LjuvernicaSweM",
               "LrealiCatF", "LrealiCatM",
               "LsinapisSweF", "LsinapisSweM",
               "LsinapisAstM", "LsinapisCatF", "LsinapisCatM")


#List of samples for external tip EBRs
species<-c("LjuvernicaSwe", "LrealiCat", "LsinapisSwe", "LsinapisCat")

header_F<-c("CHROM_F", "START_F", "STOP_F", "ORIENTATION_F", "FILE_F")
header_M<-c("CHROM_M", "START_M", "STOP_M", "ORIENTATION_M", "FILE_M")

ref<-read.table("BUSCO_basic/geneLists/genes.LsinapisAstM.busco.list")
colnames(ref)<-c("CHROM", "START_R", "STOP_R", "ORIENTATION_R", "FILE_R")
ref$buscoID<-str_split(ref$FILE_R, pattern="_", simplify = T )[,2] #Extract busco gene names

#Read and prepare chromosome names for the Asturian reference 
AST_chrm<-read.csv("./Asturian_DToL_chr_name_map.txt", header=F, sep=" ") #Add short chromosome names
AST_chrm$short<-paste0("Chr_", AST_chrm$V3)
AST_chrm$V2<-NULL
AST_chrm$V3<-NULL
colnames(AST_chrm)<-c("CHROM", "CHROM_R")
ref_chr<-merge(ref, AST_chrm, by="CHROM")
ref_chr$CHROM<-NULL


#Read lengths of the Asturian reference genome
len<-read.table("./Astrurian_DToL_chr_length.txt")
len$Z<-str_replace_all(len$V1, "Chr_48", "Chr_Z")


#Read the data for the phylogeny for calculating rates of EBRs across branches
tree<-read.tree("../Busco_phylo/run5/Reroot.Leptidea.mafft.0.2_trimal.top1percentremoved.contree")
```


#BUSCOS present across all assemblies
make a list of all buscos present in all assemblies for dowinstream filterning. 

```{r}

for( i in 1:length(all_species)){
  busco_tmp <- read.table(paste0("BUSCO_basic/geneLists/genes.", all_species[i], ".busco.list"))
  
  busco_tmp$buscoID<-str_split(busco_tmp$V5, pattern="_", simplify = T )[,2]
  
  busco_tmp[,c(1,2,3,4,5)] <-NULL
  
  busco_list<-inner_join(busco_list, busco_tmp)
}

```


#EBRs for external tip combinations
```{r}

#dataframes
bp_fis_FMR<-data.frame()
bp_fus_FMR<-data.frame()
#block_FMR_df2<-data.frame()
block_FMR_df2_filt<-data.frame()

for(i in 1:length(species)){
  
  sF<-read.table(paste0("BUSCO_basic/geneLists/genes.", species[i],"F.busco.list"))
  colnames(sF)<-header_F
  sF$buscoID<-str_split(sF$FILE_F, pattern="_", simplify = T )[,2]
   if(length(unique(duplicated(sF$buscoID))) > 1){
    print("Duplicated BUSCO IDs")
    break
  }
  
  sM<-read.table(paste0("BUSCO_basic/geneLists/genes.", species[i],"M.busco.list"))
  colnames(sM)<-header_M
  sM$buscoID<-str_split(sM$FILE_M, pattern="_", simplify = T )[,2]
  if(length(unique(duplicated(sM$buscoID))) > 1){
    print("Duplicated BUSCO IDs")
    break
  }
  
  #Include only BUSCOS present in both 
  shared_tmp<-c(sM$buscoID, sF$buscoID)
  
  shared_FM<-shared_tmp[duplicated(shared_tmp)]
  
  shared_FMR_tmp<-c(shared_FM, ref_chr$buscoID)
  
  shared_FMR<-shared_FMR_tmp[duplicated(shared_FMR_tmp)]
  
  sF_shared<-sF[sF$buscoID %in% shared_FMR,]
  sM_shared<-sM[sM$buscoID %in% shared_FMR,]
  merge_FM<-merge(sF_shared,sM_shared, by="buscoID")
  
  #Add the ref
  ref_shared<-ref_chr[ref_chr$buscoID %in% shared_FMR, ]
  merge_FMR<-merge(merge_FM, ref_shared, by="buscoID")
  
  merge_FMR_ord<-merge_FMR[,c(1,2,7,16,3,4,8,9,12,13,5,10,14,6,11,15)] 
  
  #First assign blocks based on pairwise comparison of F and M scafs
  merge_FMR_arr<-merge_FMR_ord[merge_FMR_ord$buscoID %in% busco_list$buscoID,] %>% #Filter for busco ids present in all species here.
  arrange(CHROM_F, START_F)

  n=1 #keep
  k=1#keep
  block_FMR<-data.frame()
    for(m in 1:length(merge_FMR_arr$CHROM_F)){
    
    if(m+1 > length(merge_FMR_arr$CHROM_F)){
      if(block_FMR$CHROM_M[m-1] == merge_FMR_arr$CHROM_M[m] ){
        block_FMR<-rbind(block_FMR,cbind(lndf1, block=k ) )
      }else{
        k=k+1
        block_FMR<-rbind(block_FMR,cbind(lndf1, block=k ) )
      }
    }else{
      lndf1<-merge_FMR_arr[m,]
      lndf2<-merge_FMR_arr[m+1,]
      if(lndf1$CHROM_F == lndf2$CHROM_F & lndf1$CHROM_M == lndf2$CHROM_M){
        
        block_FMR<-rbind(block_FMR,cbind(lndf1, block=k ) )
        
      }else{
       
        block_FMR<-rbind(block_FMR,cbind(lndf1, block=k ) )
        k=k+1
      }
    }
    }
  
  #Remove blocks >5 Buscos
  filt_FMR2 <- block_FMR %>%
  group_by(block) %>%
  filter(n() >= 5) %>%
  mutate(BlockStart_R = min(START_R),
     BlockStop_R = max(STOP_R),
     BlockStart_F = min(START_F),#
     BlockStop_F = max(STOP_F),#
     BlockStart_M = min(START_M),#
     BlockStop_M = max(STOP_M)) %>%
  ungroup()
  
  #Make the look up table to be used to assign EBR ranges based on the direction of the buscos in 
  lookup<-data.frame()
  for(u in 1:length(unique(filt_FMR2$CHROM_R))){
    C<-filt_FMR2[filt_FMR2$CHROM_R == unique(filt_FMR2$CHROM_R)[u],]
    Cl<-len[len$Z == unique(filt_FMR2$CHROM_R)[u],]
    
    ordC_tmp<- C %>% 
      arrange(START_R)
    
    ordC<-ordC_tmp[,c(9,10,13,16,1,4)]
    
    ordCl<-rbind(cbind(START_R=0, STOP_R=0, ORIENTATION_R=0, FILE_R="START", buscoID="START",CHROM_R=unique(filt_FMR2$CHROM_R)[u]), 
            ordC,
            cbind(START_R=Cl$V2, STOP_R=Cl$V2, ORIENTATION_R=0, FILE_R="STOP", buscoID="STOP",CHROM_R=unique(filt_FMR2$CHROM_R)[u]))
    
    
    
    ordCl$BUSCO_ORDER <- 1:nrow(ordCl)

    lookup<-rbind(lookup, ordCl)
  }
  
   
  #Assign blocks based on the Reference coordinate system. 
  block_FMR_tmp_arr<-filt_FMR2 %>%
  arrange(CHROM_R, START_R)

  n=1 #keep
  k=1#keep
  block_FMR_R<-data.frame()
    for(m in 1:length(block_FMR_tmp_arr$CHROM_R)){
    
    if(m+1 > length(block_FMR_tmp_arr$CHROM_R)){
      if(block_FMR_tmp_arr$CHROM_R[m-1] == block_FMR_tmp_arr$CHROM_R[m] & block_FMR_tmp_arr$block[m-1] == block_FMR_tmp_arr$block[m]){
        block_FMR_R<-rbind(block_FMR_R,cbind(lndf1, block_R=k ) )
      }else{
        k=k+1
        block_FMR_R<-rbind(block_FMR_R,cbind(lndf1, block_R=k ) )
      }
    }else{
      lndf1<-block_FMR_tmp_arr[m,]
      lndf2<-block_FMR_tmp_arr[m+1,]
      if(lndf1$CHROM_R == lndf2$CHROM_R & lndf1$block == lndf2$block){
        
        block_FMR_R<-rbind(block_FMR_R,cbind(lndf1, block_R=k ) )
        
      }else{
       
        block_FMR_R<-rbind(block_FMR_R,cbind(lndf1, block_R=k ) )
        k=k+1
      }
    }
    }    
    
    #Add start and stop of blocks relative to the reference coordinate system
    filt_FMR_R <- block_FMR_R %>%
      group_by(block_R) %>%
      mutate(BlockStart_Rvis = min(START_R),
             BlockStop_Rvis = max(STOP_R)) %>%
      ungroup()

    block_FMR_df2_filt<-rbind(block_FMR_df2_filt, filt_FMR_R)
    
    
    #Extract EBRs 
    
      for(e in 1:length(unique(filt_FMR2$CHROM_M))){
        
        bptmp1<-filt_FMR2[filt_FMR2$CHROM_M == unique(filt_FMR2$CHROM_M)[e],]


        if(length(paste0(unique(bptmp1$block),":", unique(bptmp1$CHROM_M), ":", unique(bptmp1$CHROM_F))) > 1){
          
          
          bp1<- bptmp1 %>% 
            arrange(CHROM_M, START_M)

          #Different in DToL same in CAR
          for(r in 1:length(bp1$CHROM_F)){

            if(r+1 > length(bp1$CHROM_F) ){

            }else{
              if(bp1[r,]$CHROM_F != bp1[r+1,]$CHROM_F ){
                    
                
                  #First end              
                  BuscoStart_R=bp1[r,]$START_R
                  BuscoStop_R=bp1[r,]$STOP_R
                  BuscoStart_F=bp1[r,]$START_F
                  BuscoStop_F=bp1[r,]$STOP_F
                  BuscoStart_M=bp1[r,]$START_M
                  BuscoStop_M=bp1[r,]$STOP_M
                  BuscoID1=bp1[r,]$buscoID
                  BuscoID2=bp1[r+1,]$buscoID
                  BuscoIDR1=bp1[r,]$buscoID
                  line="r"
                  
                  if(bp1[r,]$CHROM_R != bp1[r-1,]$CHROM_R ){
                    print("No previous BUSCO 1")
                  }else{
                  
                    #For 5` to 3` order in ref
                    if(bp1[r-1,]$START_R < bp1[r,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp1[r,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                        BPStopR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_R
                        BPStartR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$STOP_R
                      
                    }
                    
                    #For 3` to 5` order in ref
                    if(bp1[r-1,]$START_R > bp1[r,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp1[r,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                        BPStartR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_R
                        BPStopR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$START_R
                      
                    }
  
                    
                    bp_fis_FMR<-rbind(bp_fis_FMR, cbind(bp1[r,], Fis=paste0("Fis",n),
                                                        BuscoStart_R,BuscoStop_R,
                                                        BuscoStart_F,BuscoStop_F,
                                                        BuscoStart_M,BuscoStop_M,
                                                        BuscoID1,BuscoID2,
                                                        BuscoIDR1,BuscoIDR2,
                                                        BPStartR,BPStopR,
                                                        line))
                 
                    
                  }
  
                  #Other end
                  
                  BuscoStart_R=bp1[r+1,]$START_R
                  BuscoStop_R=bp1[r+1,]$STOP_R
                  BuscoStart_F=bp1[r+1,]$START_F
                  BuscoStop_F=bp1[r+1,]$STOP_F
                  BuscoStart_M=bp1[r+1,]$START_M
                  BuscoStop_M=bp1[r+1,]$STOP_M
                  BuscoID1=bp1[r,]$buscoID
                  BuscoID2=bp1[r+1,]$buscoID
                  BuscoIDR1=bp1[r+1,]$buscoID
                  line="r+1"
                  
                  if(bp1[r+1,]$CHROM_R != bp1[r+2,]$CHROM_R ){
                    print("No following BUSCO 2")
                  }else{
                  
                    #For 5` to 3` order in ref
                    if(bp1[r+1,]$START_R < bp1[r+2,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp1[r+1,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                        BPStartR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_R
                        BPStopR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$START_R
                      
                    }
                    
                    #For 3` to 5` order in ref
                    if(bp1[r+1,]$START_R > bp1[r+2,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp1[r+1,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                        BPStopR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_R
                        BPStartR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$STOP_R
                      
                    }
  
                    
                    bp_fis_FMR<-rbind(bp_fis_FMR, cbind(bp1[r+1,], Fis=paste0("Fis",n),
                                                        BuscoStart_R,BuscoStop_R,
                                                        BuscoStart_F,BuscoStop_F,
                                                        BuscoStart_M,BuscoStop_M,
                                                        BuscoID1,BuscoID2,
                                                        BuscoIDR1,BuscoIDR2,
                                                        BPStartR,BPStopR,
                                                        line))
                    n=n+1
                    
                  }
  
                                  
            }else{
            }
              }
            }
      }

      }
    
          for(e in 1:length(unique(filt_FMR2$CHROM_F))){
        
        bptmp2<-filt_FMR2[filt_FMR2$CHROM_F == unique(filt_FMR2$CHROM_F)[e],]
        
        if(length(paste0(unique(bptmp2$block),":", unique(bptmp2$CHROM_M), ":", unique(bptmp2$CHROM_F))) > 1){
        
          bp2<- bptmp2 %>% 
            arrange(CHROM_F, START_F)
          #Different in CAR same DToL

          for(r in 1:length(bp2$CHROM_M)){


            if(r+1 > length(bp2$CHROM_M) ){

            }else{

              if(bp2[r,]$CHROM_M != bp2[r+1,]$CHROM_M ){
                
                               
                  #First end              
                  BuscoStart_R=bp2[r,]$START_R
                  BuscoStop_R=bp2[r,]$STOP_R
                  BuscoStart_F=bp2[r,]$START_F
                  BuscoStop_F=bp2[r,]$STOP_F
                  BuscoStart_M=bp2[r,]$START_M
                  BuscoStop_M=bp2[r,]$STOP_M
                  BuscoID1=bp2[r,]$buscoID
                  BuscoID2=bp2[r+1,]$buscoID
                  BuscoIDR1=bp2[r,]$buscoID
                  line="r"
                  
                  if(bp2[r,]$CHROM_R != bp2[r-1,]$CHROM_R ){
                    print("No previous BUSCO 3")
                  }else{
                  
                    #For 5` to 3` order in ref
                    if(bp2[r-1,]$START_R < bp2[r,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp2[r,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                        BPStopR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_R
                        BPStartR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$STOP_R
                      
                    }
                    
                    #For 3` to 5` order in ref
                    if(bp2[r-1,]$START_R > bp2[r,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp2[r,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                        BPStartR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_R
                        BPStopR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$START_R
                      
                    }
  
                    
                    bp_fus_FMR<-rbind(bp_fus_FMR, cbind(bp2[r,], Fus=paste0("Fus",n),
                                                        BuscoStart_R,BuscoStop_R,
                                                        BuscoStart_F,BuscoStop_F,
                                                        BuscoStart_M,BuscoStop_M,
                                                        BuscoID1,BuscoID2,
                                                        BuscoIDR1,BuscoIDR2,
                                                        BPStartR,BPStopR,
                                                        line))
                    
                    
                  }
  
                  #Other end
                  
                  BuscoStart_R=bp2[r+1,]$START_R
                  BuscoStop_R=bp2[r+1,]$STOP_R
                  BuscoStart_F=bp2[r+1,]$START_F
                  BuscoStop_F=bp2[r+1,]$STOP_F
                  BuscoStart_M=bp2[r+1,]$START_M
                  BuscoStop_M=bp2[r+1,]$STOP_M
                  BuscoID1=bp2[r,]$buscoID
                  BuscoID2=bp2[r+1,]$buscoID
                  BuscoIDR1=bp2[r+1,]$buscoID
                  line="r+1"
                  
                  if(bp2[r+1,]$CHROM_R != bp2[r+2,]$CHROM_R ){
                    print(paste0("No following BUSCO 4 ", bp2[r+1,]$CHROM_R ," ", bp2[r,]$CHROM_R))
                    if(bp2[r,]$CHROM_R != bp2[r+1,]$CHROM_R ){
                      print("N0 previous BUSCO 5")
                    }else{
                      #For 5` to 3` order in ref
                        if(bp2[r,]$START_R < bp2[r+1,]$START_R){
                          
                            cur_chr<-lookup[lookup$CHROM_R == bp2[r+1,]$CHROM_R,]
                            
                            cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                          
                            BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                            BPStartR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_R
                            BPStopR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$START_R
                          
                        }
                        
                        #For 3` to 5` order in ref
                        if(bp2[r,]$START_R > bp2[r+1,]$START_R){
                          
                            cur_chr<-lookup[lookup$CHROM_R == bp2[r+1,]$CHROM_R,]
                            
                            cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                          
                            BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                            BPStopR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_R
                            BPStartR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$STOP_R
                          
                        }
      
                        
                        bp_fus_FMR<-rbind(bp_fus_FMR, cbind(bp2[r+1,], Fus=paste0("Fus",n),
                                                            BuscoStart_R,BuscoStop_R,
                                                            BuscoStart_F,BuscoStop_F,
                                                            BuscoStart_M,BuscoStop_M,
                                                            BuscoID1,BuscoID2,
                                                            BuscoIDR1,BuscoIDR2,
                                                            BPStartR,BPStopR,
                                                            line))
                        n=n+1
                        
                   }
                  }else{
                  
                    #For 5` to 3` order in ref
                    if(bp2[r+1,]$START_R < bp2[r+2,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp2[r+1,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                        BPStartR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_R
                        BPStopR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$START_R
                      
                    }
                    
                    #For 3` to 5` order in ref
                    if(bp2[r+1,]$START_R > bp2[r+2,]$START_R){
                      
                        cur_chr<-lookup[lookup$CHROM_R == bp2[r+1,]$CHROM_R,]
                        
                        cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$BUSCO_ORDER
                      
                        BuscoIDR2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                        BPStopR=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_R
                        BPStartR=cur_chr[cur_chr$buscoID ==  BuscoIDR1, ]$STOP_R
                      
                    }
  
                    
                    bp_fus_FMR<-rbind(bp_fus_FMR, cbind(bp2[r+1,], Fus=paste0("Fus",n),
                                                        BuscoStart_R,BuscoStop_R,
                                                        BuscoStart_F,BuscoStop_F,
                                                        BuscoStart_M,BuscoStop_M,
                                                        BuscoID1,BuscoID2,
                                                        BuscoIDR1,BuscoIDR2,
                                                        BPStartR,BPStopR,
                                                        line))
                    n=n+1
                    
                  }
              }else{


              }
            }
            }
          }
      }
    
}


#Assign the NODE label based on sample id matching based on Phylogeny

block_FMR_df2_filt$sF<-str_split(block_FMR_df2_filt$FILE_F,".busco", simplify = T )[,1]
block_FMR_df2_filt$sM<-str_split(block_FMR_df2_filt$FILE_M,".busco", simplify = T )[,1]
block_FMR_df2_filt$sR<-str_split(block_FMR_df2_filt$FILE_R,".busco", simplify = T )[,1]
block_FMR_df2_filt<- block_FMR_df2_filt %>% 
              mutate(NODE = ifelse(sF == "LsinapisSweF", "N5", 
                                   ifelse(sF == "LjuvernicaSweF", "N2", 
                                          ifelse(sF == "LrealiCatF", "N4",
                                                 ifelse(sF == "LsinapisCatF", "N7", NA)))))


bp_fis_FMR$sF<-str_split(bp_fis_FMR$FILE_F,".busco", simplify = T )[,1]
bp_fis_FMR$sM<-str_split(bp_fis_FMR$FILE_M,".busco", simplify = T )[,1]
bp_fus_FMR$sF<-str_split(bp_fus_FMR$FILE_F,".busco", simplify = T )[,1]
bp_fus_FMR$sM<-str_split(bp_fus_FMR$FILE_M,".busco", simplify = T )[,1]

bp_fis_FMR <- bp_fis_FMR %>% 
              mutate(NODE = ifelse(sF == "LsinapisSweF", "N5", ifelse(sF == "LjuvernicaSweF", "N2", NA)))

bp_fus_FMR <- bp_fus_FMR %>% 
              mutate(NODE = ifelse(sF == "LsinapisSweF", "N5", ifelse(sF == "LjuvernicaSweF", "N2", NA)))


#Check that EBR start position is smaller than EBR stop position
as.numeric(bp_fis_FMR$BPStartR) < as.numeric(bp_fis_FMR$BPStopR)
as.numeric(bp_fus_FMR$BPStartR) < as.numeric(bp_fus_FMR$BPStopR)

```

#Agora ortholists preparation
Busco gene mapping to the Asturian reference system 
```{r}

OrthoList_folder <-  "BUSCO_basic/withData"
OrthoList_fussion <- list.files(OrthoList_folder, full.names = TRUE, pattern = "OrthoList_N")

#grep for all samples execept the reference
samples<-OrthoList_fussion[!grepl("AstM",OrthoList_fussion)]

new_coords<-data.frame()
for(l in 1:length(samples)){

  ref= OrthoList_fussion[grepl(paste0(unlist(str_split(samples[l], "_"))[3],"_LsinapisAstM"),OrthoList_fussion)] 
  sam=samples[l] 
  
  if(length(ref)==0){
    
    print("No NODE support for ref")
    }else{
  
    data_ref<-read.csv(ref, sep = "\t", header=F)
    colnames(data_ref)<-c("CHROM_DToL", "START_DToL", "STOP_DToL", "ORIENTATION_DToL", "FILE_DToL", "ANC_CHROM_DToL", "ANC_START_DToL", "ANC_STOP_DToL", "ANC_ORIENTATION_DToL", "BUSCO")
    
    data_ref<-data_ref[!grepl("singleton", data_ref$ANC_CHROM_DToL ),]
    
    data_ref2<-data.frame()
    for(i in 1:length(data_ref$ANC_START_DToL)){
      if(data_ref$ANC_START_DToL[i]<=data_ref$ANC_STOP_DToL[i]){
        data_ref2<-rbind(data_ref2, cbind(join=paste0(data_ref$ANC_CHROM_DToL[i],":",data_ref$ANC_START_DToL[i], "-",data_ref$ANC_STOP_DToL[i]), data_ref[i,]))
      }else{
        data_ref2<-rbind(data_ref2, cbind(join=paste0(data_ref$ANC_CHROM_DToL[i],":",data_ref$ANC_STOP_DToL[i], "-",data_ref$ANC_START_DToL[i]), data_ref[i,]))
      }
    }
    
    data<-read.csv(sam, sep = "\t", header=F)
    colnames(data)<-c("CHROM", "START", "STOP", "ORIENTATION", "FILE", "ANC_CHROM", "ANC_START", "ANC_STOP", "ANC_ORIENTATION", "BUSCO")
    
    data<-data[!grepl("singleton", data$ANC_CHROM ),]
    
    data2<-data.frame()
    for(i in 1:length(data$ANC_START)){
      if(data$ANC_START[i]<=data$ANC_STOP[i]){
        data2<-rbind(data2, cbind(join=paste0(data$ANC_CHROM[i],":",data$ANC_START[i], "-",data$ANC_STOP[i]), data[i,]))
      }else{
        data2<-rbind(data_2, cbind(join=paste0(data$ANC_CHROM[i],":",data$ANC_STOP[i], "-",data$ANC_START[i]), data[i,]))
      }
    }
    
    data2$SPECIES<-unlist(str_split(sam, pattern="_"))[4]
    
    dr2<-data_ref2[!duplicated(data_ref2$join) & !duplicated(data_ref2$BUSCO),]
    d2<-data2[!duplicated(data2$join) & !duplicated(data2$BUSCO),]
    
    data_join<-inner_join(x = dr2, y = d2, by = c( "join","BUSCO"))
    
    identical(data_join$ANC_CHROM_DToL,data_join$ANC_CHROM_DToL)
    identical(data_join$ANC_START_DToL,data_join$ANC_START_DToL)
    identical(data_join$ANC_STOP_DToL,data_join$ANC_STOP_DToL)
    
    data_join$ANC_CHROM_DToL<-NULL
    data_join$ANC_START_DToL<-NULL
    data_join$ANC_STOP_DToL<-NULL
    
    #Rename chromos
    AST_chrm<-read.csv("./Asturian_DToL_chr_name_map.txt", header=F, sep=" ")
    
    AST_chrm$short<-paste0("Chr_", AST_chrm$V3)
    
    AST_chrm$V2<-NULL
    AST_chrm$V3<-NULL
    colnames(AST_chrm)<-c("CHROM_DToL", "short")
    
    
    data_join<-merge(data_join, AST_chrm, by="CHROM_DToL")
    
    data_join_ord<-data_join[,c(17,19,3,4,5,12,14,15,16,18,8,9,10,11,6,13)] 
    colnames(data_join_ord)[2]<-"CHROM_DToL"
    
    data_join_ord %>% 
      arrange(CHROM_DToL, START_DToL)
    
    data_join_ord %>% 
      arrange(CHROM, START)
    
    #ALWAYS START smaller than STOP
    unique(data_join_ord$START_DToL <= data_join_ord$STOP_DToL)
    unique(data_join_ord$START <= data_join_ord$STOP)
    unique(data_join_ord$ANC_START <= data_join_ord$ANC_STOP)
    
    data_join_ord$NODE<-unlist(str_split(samples[l], "_"))[3]
    new_coords<-rbind(new_coords,data_join_ord )
    
    
    #structure the data for the reference as well

    new_coords_ref<-merge(data_ref2, AST_chrm, by="CHROM_DToL")
    new_coords_ref$SPECIES<-unlist(str_split(data_ref2$FILE_DToL, ".busco" ))[1]
    new_coords_ref$CHROM_DToL<-NULL
    new_coords_ref$join<-NULL
    new_coords_ref$CHROM<-new_coords_ref$short
    new_coords_ref$START<-new_coords_ref$START_DToL
    new_coords_ref$STOP<-new_coords_ref$STOP_DToL
    new_coords_ref$ORIENTATION<-new_coords_ref$ORIENTATION_DToL
    new_coords_ref$NODE<-unlist(str_split(samples[l], "_"))[3]
    new_coords_ref$FILE<-new_coords_ref$FILE_DToL

    
    new_coords_ref_ord<-new_coords_ref[,c(11,10,1,2,3,5,6,7,8,9,12,13,14,15,16,4,17)]
    
    
    colnames(new_coords_ref_ord)<-c("SPECIES","CHROM_DToL","START_DToL","STOP_DToL","ORIENTATION_DToL","ANC_CHROM","ANC_START","ANC_STOP","ANC_ORIENTATION","BUSCO","CHROM","START","STOP","ORIENTATION","NODE","FILE_DToL","FILE")  
    
    
    new_coords<-rbind(new_coords,new_coords_ref_ord )
    
    new_coords<-new_coords[!duplicated(new_coords),]
    
    }
}

unique(new_coords$SPECIES)
unique(new_coords$NODE)

View(new_coords)

i=1
for(i in 1:length(unique(new_coords$SPECIES))){
  
  sdf<-new_coords[new_coords$SPECIES == unique(new_coords$SPECIES)[i],]
  
  j=1
  for(j in 1:length(unique(sdf$NODE))){
    ndf<-sdf[sdf$NODE == unique(sdf$NODE)[j],]
    
    p=ggplot(ndf, aes(xmin=START_DToL, xmax=STOP_DToL, y=CHROM_DToL, col=ANC_CHROM))+
      geom_linerange(lwd=2)+
      guides(col="none")+
      theme_bw()
    
    print(p)
    }
  
}

#Write buscoIDs in the same format as the Straight buscos. 

Agora_data <- new_coords %>% 
  mutate(buscoID = str_split(FILE, "_", simplify = T)[,2]) 


#Check for duplicated busco ids, sanity check
for(i in 1:length(unique(Agora_data$SPECIES))){
  
  sdf<-Agora_data[Agora_data$SPECIES == unique(Agora_data$SPECIES)[i],]
  
  j=1
  for(j in 1:length(gtools::mixedsort(unique(sdf$NODE)))){
    ndf<-sdf[sdf$NODE == gtools::mixedsort(unique(sdf$NODE))[j],]
    
     if(length(unique(duplicated(ndf$buscoID))) > 1){
      print(paste0(unique(ndf$NODE),"Duplicated BUSCO IDs for:", unique(ndf$SPECIES)))
      break
     }else{
      print(paste0(unique(ndf$NODE),"; No Duplicated BUSCO IDs for:", unique(ndf$SPECIES)))
  }
    }
  
}



```


#EBR identification for AGORA

```{r}

block_df1<-data.frame() #Blocks background

bp_fis<-data.frame()
bp_fus<-data.frame()
#per species identification of EBRs
for(i in 1:length(unique(Agora_data$SPECIES))){
  
  sdf<-Agora_data[Agora_data$SPECIES == unique(Agora_data$SPECIES)[i],]
  
  #per node 
  n=1 #keep
  for(j in 1:length(unique(sdf$NODE))){
    ndf<-sdf[sdf$NODE == unique(sdf$NODE)[j],]
    
    #MAKE BLOCKS
    #BLOCKS needs to be based on REF coordinate system but FILTERED <5 BUSCOS based on pairwise comparison of CAR and SPECIES scafs. Breakpoints still have to be found based on pairwise comparison of CAR and SPECIES scafs
  
    #First assign blocks based on pairwise comparison of CAR and SPECIES scafs
    ndf_arr<-ndf[ndf$buscoID %in% busco_list$buscoID, ] %>% #Remove buscos not present in all species
    arrange(CHROM, START)
  
    n=1 #keep
    k=1#keep
    blocks<-data.frame()
      for(m in 1:length(ndf_arr$CHROM)){
      
      if(m+1 > length(ndf_arr$CHROM)){
        if(ndf_arr$ANC_CHROM[m-1] == ndf_arr$ANC_CHROM[m] ){
          blocks<-rbind(blocks,cbind(lndf1, block=k ) )
        }else{
          k=k+1
          blocks_CHROM<-rbind(blocks_CHROM,cbind(lndf1, block=k ) )
        }
      }else{
        lndf1<-ndf_arr[m,]
        lndf2<-ndf_arr[m+1,]
        if(lndf1$CHROM == lndf2$CHROM & lndf1$ANC_CHROM == lndf2$ANC_CHROM){
          
          blocks<-rbind(blocks,cbind(lndf1, block=k ) )
          
        }else{
         
          blocks<-rbind(blocks,cbind(lndf1, block=k ) )
          k=k+1
        }
      }
      }
    
    
    
    #Remove blocks >3 Buscos
    #This limit is lower because the ancestral karotype models are implemented with AGORA
    filt2 <- blocks %>%
    group_by(block) %>%
    filter(n() >= 3) %>%
    mutate(BlockStart_R = min(START_DToL),
       BlockStop_R = max(STOP_DToL),
       BlockStart_CHROM = min(START),
       BlockStop_CHROM = max(STOP),
       BlockStart_ANC = min(ANC_START),
       BlockStop_ANC = max(ANC_STOP)) %>%
    ungroup()
      
    block_df1<-rbind(block_df1, filt2)
    
    #MAKE LOOK UP TABLE 
    len<-read.table("../DELLY/DToL.length")
    len$Z<-str_replace_all(len$V1, "Chr_48", "Chr_Z")
    
    lookup<-data.frame()
    for(u in 1:length(unique(filt2$CHROM_DToL))){
      C<-filt2[filt2$CHROM_DToL == unique(filt2$CHROM_DToL)[u],]
      Cl<-len[len$Z == unique(filt2$CHROM_DToL)[u],]
      
      ordC_tmp<- C %>% 
        arrange(START_DToL)
      
      ordC<-ordC_tmp[,c(3,4,5,15,18,2)]
      
      ordCl<-rbind(cbind(START_DToL=0, STOP_DToL=0, ORIENTATION_DToL=0, FILE_DToL="START", buscoID="START", CHROM_DToL=unique(filt2$CHROM_DToL)[u]), 
              ordC,
              cbind(START_DToL=Cl$V2, STOP_DToL=Cl$V2, ORIENTATION_DToL=0, FILE_DToL="STOP", buscoID="STOP", CHROM_DToL=unique(filt2$CHROM_DToL)[u]))
      
      
      
      ordCl$BUSCO_ORDER <- 1:nrow(ordCl)
  
      lookup<-rbind(lookup, ordCl)
    }
    
    #Extract EBRs
    #First loop deals with Fissions
    for(e in 1:length(unique(filt2$ANC_CHROM))){
      
      bp1_tmp<-filt2[filt2$ANC_CHROM == unique(filt2$ANC_CHROM)[e],]


      if(length(paste0(unique(bp1_tmp$block),":", unique(bp1_tmp$ANC_CHROM), ":", unique(bp1_tmp$CHROM))) > 1){
        

        bp1<- bp1_tmp %>% 
          arrange(ANC_CHROM, ANC_START)

        #Different in DToL same in CAR
        for(r in 1:length(bp1$CHROM)){

          if(r+2 > length(bp1$CHROM) | r-1 == 0 ){ #Excludes blocks of a single busco at the end of alignments

          }else{

            if(bp1[r,]$CHROM != bp1[r+1,]$CHROM ){
                  
              #First end              
              BuscoStart_DToL=bp1[r,]$START_DToL
              BuscoStop_DToL=bp1[r,]$STOP_DToL
              BuscoStart_CHROM=bp1[r,]$START
              BuscoStop_CHROM=bp1[r,]$STOP
              BuscoStart_ANC=bp1[r,]$ANC_START
              BuscoStop_ANC=bp1[r,]$ANC_STOP
              BuscoID1=bp1[r,]$buscoID
              BuscoID2=bp1[r+1,]$buscoID
              BuscoIDDToL1=bp1[r,]$buscoID
              line="r"
              
              if(bp1[r,]$CHROM_DToL != bp1[r-1,]$CHROM_DToL){
                print(paste0("First loop: No previous BUSCO ",e, " Species:", unique(Agora_data$SPECIES)[i]))
                
                if(bp1[r,]$CHROM_DToL != bp1[r+1,]$CHROM_DToL){
                  print(paste0("First loop:  No following BUSCO  ",e, " Species:", unique(Agora_data$SPECIES)[i]))
                }
                 
              }else{
              
                #For 5` to 3` order in ref
                if(bp1[r-1,]$START_DToL < bp1[r,]$START_DToL){
                  
                    cur_chr<-lookup[lookup$CHROM_DToL == bp1[r,]$CHROM_DToL,]
                    
                    cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                  
                    BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                    BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_DToL)
                    BPStartDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$STOP_DToL)
                    
                    if(BPStopDToL < BPStartDToL){
                      BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+2),]$START_DToL)
                      
                    }
                  
                }
                
                #For 3` to 5` order in ref
                if(bp1[r-1,]$START_DToL > bp1[r,]$START_DToL){
                  
                    cur_chr<-lookup[lookup$CHROM_DToL == bp1[r,]$CHROM_DToL,]
                    
                    cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                  
                    BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                    BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_DToL)
                    BPStopDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$START_DToL)
                    
                    if(BPStopDToL < BPStartDToL){
                      BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-2),]$STOP_DToL)
                    }
                  
                }

                
                bp_fis<-rbind(bp_fis, cbind(bp1[r,], Fis=paste0("Fis",n),
                                                    BuscoStart_DToL,BuscoStop_DToL,
                                                    BuscoStart_CHROM,BuscoStop_CHROM,
                                                    BuscoStart_ANC,BuscoStop_ANC,
                                                    BuscoID1,BuscoID2,
                                                    BuscoIDDToL1,BuscoIDDToL2,
                                                    BPStartDToL,BPStopDToL,
                                                    line))
              
                
              }

              #Other end
              
              BuscoStart_DToL=bp1[r+1,]$START_DToL
              BuscoStop_DToL=bp1[r+1,]$STOP_DToL
              BuscoStart_CHROM=bp1[r+1,]$START
              BuscoStop_CHROM=bp1[r+1,]$STOP
              BuscoStart_ANC=bp1[r+1,]$ANC_START
              BuscoStop_ANC=bp1[r+1,]$ANC_STOP
              BuscoID1=bp1[r,]$buscoID
              BuscoID2=bp1[r+1,]$buscoID
              BuscoIDDToL1=bp1[r+1,]$buscoID
              line="r+1"
              
              if(bp1[r+1,]$CHROM_DToL != bp1[r+2,]$CHROM_DToL ){
                print(paste0("First loop: No following BUSCO ",e, " Species:", unique(Agora_data$SPECIES)[i]))
                
                if(bp1[r+1,]$CHROM_DToL != bp1[r,]$CHROM_DToL){
                  print(paste0("First loop: No previous BUSCO ",e, " Species:", unique(Agora_data$SPECIES)[i]))
                }
                
                n=n+1
              }else{
              
                #For 5` to 3` order in ref
                if(bp1[r+1,]$START_DToL < bp1[r+2,]$START_DToL){
                  
                    cur_chr<-lookup[lookup$CHROM_DToL == bp1[r+1,]$CHROM_DToL,]
                    
                    cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                  
                    BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                    BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_DToL)
                    BPStopDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$START_DToL)
                  
                    if(BPStopDToL < BPStartDToL){
                      BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-2),]$STOP_DToL)
                    }
                    
                }
                
                #For 3` to 5` order in ref
                if(bp1[r+1,]$START_DToL > bp1[r+2,]$START_DToL){
                  
                    cur_chr<-lookup[lookup$CHROM_DToL == bp1[r+1,]$CHROM_DToL,]
                    
                    cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                  
                    BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                    BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_DToL)
                    BPStartDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$STOP_DToL)
                    
                    if(BPStopDToL < BPStartDToL){
                      BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+2),]$START_DToL)
                    }
                  
                }

                
                bp_fis<-rbind(bp_fis, cbind(bp1[r+1,], Fis=paste0("Fis",n),
                                                    BuscoStart_DToL,BuscoStop_DToL,
                                                    BuscoStart_CHROM,BuscoStop_CHROM,
                                                    BuscoStart_ANC,BuscoStop_ANC,
                                                    BuscoID1,BuscoID2,
                                                    BuscoIDDToL1,BuscoIDDToL2,
                                                    BPStartDToL,BPStopDToL,
                                                    line))
                n=n+1
                
              }

            }else{

            }
          }
            }
          }
    }
  
    #This loop assigns Fussions
    for(e in 1:length(unique(filt2$CHROM))){
      
      bp2_tmp<-filt2[filt2$CHROM == unique(filt2$CHROM)[e],]
      
      if(length(paste0(unique(bp2_tmp$block),":", unique(bp2_tmp$ANC_CHROM), ":", unique(bp2_tmp$CHROM))) > 1){
        
        bp2<- bp2_tmp %>% 
          arrange(CHROM, START)
        #Different in CAR same DToL
        r=1
        for(r in 1:length(bp2$ANC_CHROM)){


          if(r+2 > length(bp2$ANC_CHROM) | r-1 == 0  ){#Excludes blocks of a single busco at the end of alignments

          }else{
            
            if(bp2[r,]$ANC_CHROM != bp2[r+1,]$ANC_CHROM ){
            
           
            #First end              
            BuscoStart_DToL=bp2[r,]$START_DToL
            BuscoStop_DToL=bp2[r,]$STOP_DToL
            BuscoStart_CHROM=bp2[r,]$START
            BuscoStop_CHROM=bp2[r,]$STOP
            BuscoStart_ANC=bp2[r,]$ANC_START
            BuscoStop_ANC=bp2[r,]$ANC_STOP
            BuscoID1=bp2[r,]$buscoID
            BuscoID2=bp2[r+1,]$buscoID
            BuscoIDDToL1=bp2[r,]$buscoID
            line="r"
            
            if(bp2[r,]$CHROM_DToL != bp2[r-1,]$CHROM_DToL ){
              print(paste0("Second loop: No previous BUSCO ",n," ",bp2[r,]$NODE , " Species:", unique(Agora_data$SPECIES)[i]))
                
                if(bp2[r,]$CHROM_DToL != bp2[r+1,]$CHROM_DToL){
                  print(paste0("Second loop: No following BUSCO ",n," ",bp2[r,]$NODE , " Species:", unique(Agora_data$SPECIES)[i]))
                }
                
              
            }else{
            
              #For 5` to 3` order in ref
              if(bp2[r-1,]$START_DToL < bp2[r,]$START_DToL){
                
                  cur_chr<-lookup[lookup$CHROM_DToL == bp2[r,]$CHROM_DToL,]
                  
                  cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                
                  BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                  BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_DToL)
                  BPStartDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$STOP_DToL)
                  
                  if(BPStopDToL < BPStartDToL){
                    BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+2),]$START_DToL)
                  }
                  
                                  
              }
              
              #For 3` to 5` order in ref
              if(bp2[r-1,]$START_DToL > bp2[r,]$START_DToL){
                
                  cur_chr<-lookup[lookup$CHROM_DToL == bp2[r,]$CHROM_DToL,]
                  
                  cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                
                  BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                  BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_DToL)
                  BPStopDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$START_DToL)
                  
                  if(BPStopDToL < BPStartDToL){
                    BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-2),]$STOP_DToL)
                  }
                
              }

              
              bp_fus<-rbind(bp_fus, cbind(bp2[r,], Fus=paste0("Fus",n),
                                                  BuscoStart_DToL,BuscoStop_DToL,
                                                  BuscoStart_CHROM,BuscoStop_CHROM,
                                                  BuscoStart_ANC,BuscoStop_ANC,
                                                  BuscoID1,BuscoID2,
                                                  BuscoIDDToL1,BuscoIDDToL2,
                                                  BPStartDToL,BPStopDToL,
                                                  line))
            
              
            }

            #Other end
            
            BuscoStart_DToL=bp2[r+1,]$START_DToL
            BuscoStop_DToL=bp2[r+1,]$STOP_DToL
            BuscoStart_CHROM=bp2[r+1,]$START
            BuscoStop_CHROM=bp2[r+1,]$STOP
            BuscoStart_ANC=bp2[r+1,]$ANC_START
            BuscoStop_ANC=bp2[r+1,]$ANC_STOP
            BuscoID1=bp2[r,]$buscoID
            BuscoID2=bp2[r+1,]$buscoID
            BuscoIDDToL1=bp2[r+1,]$buscoID
            line="r+1"
            
            if(bp2[r+1,]$CHROM_DToL != bp2[r+2,]$CHROM_DToL ){
                print(paste0("Second loop: No following BUSCO ",n," ",bp2[r,]$NODE,  " Species:", unique(Agora_data$SPECIES)[i]))
                
                if(bp2[r+1,]$CHROM_DToL != bp2[r,]$CHROM_DToL){
                  print(paste0("Second loop:  No previous BUSCO ",n," ",bp2[r,]$NODE,  " Species:", unique(Agora_data$SPECIES)[i]))
                }
              

              n=n+1
            }else{
            
              #For 5` to 3` order in ref
              if(bp2[r+1,]$START_DToL < bp2[r+2,]$START_DToL){
                
                  cur_chr<-lookup[lookup$CHROM_DToL == bp2[r+1,]$CHROM_DToL,]
                  
                  cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                
                  BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$buscoID
                  BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-1),]$STOP_DToL)
                  BPStopDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$START_DToL)
                  
                  if(BPStopDToL < BPStartDToL){
                    BPStartDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord-2),]$STOP_DToL)
                  }
                
              }
              
              #For 3` to 5` order in ref
              if(bp2[r+1,]$START_DToL > bp2[r+2,]$START_DToL){
                
                  cur_chr<-lookup[lookup$CHROM_DToL == bp2[r+1,]$CHROM_DToL,]
                  
                  cur_ord<-cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$BUSCO_ORDER
                
                  BuscoIDDToL2=cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$buscoID
                  BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+1),]$START_DToL)
                  BPStartDToL=as.numeric(cur_chr[cur_chr$buscoID ==  BuscoIDDToL1, ]$STOP_DToL)
                  
                  
                  if(BPStopDToL < BPStartDToL){
                    BPStopDToL=as.numeric(cur_chr[cur_chr$BUSCO_ORDER == (cur_ord+2),]$START_DToL)
                  }
                
              }

              
              bp_fus<-rbind(bp_fus, cbind(bp2[r+1,], Fus=paste0("Fus",n),
                                                  BuscoStart_DToL,BuscoStop_DToL,
                                                  BuscoStart_CHROM,BuscoStop_CHROM,
                                                  BuscoStart_ANC,BuscoStop_ANC,
                                                  BuscoID1,BuscoID2,
                                                  BuscoIDDToL1,BuscoIDDToL2,
                                                  BPStartDToL,BPStopDToL,
                                                  line))
              n=n+1
              
            }   

            }else{


            }
          }
          }
        }
    }
  }
}

#Sanity checks
unique(as.numeric(bp_fis$BPStartDToL) < as.numeric(bp_fis$BPStopDToL))
unique(as.numeric(bp_fus$BPStartDToL) < as.numeric(bp_fus$BPStopDToL))
bp_fus[!as.numeric(bp_fus$BPStartDToL) < as.numeric(bp_fus$BPStopDToL),] 
```
#Filter EBRs ranges 

```{r, warning=FALSE}

header=c("CHROM","START","STOP","SPECIES","NODE","BP","BUSCO1","BUSCO2","line","ORG_CHROM")
#Same format
colnames(bp_fus)
colnames(bp_fis)
colnames(bp_fus_FMR)
colnames(bp_fis_FMR)

#subset
#agra
bp_fus_sh<-bp_fus[,c(2,37,38,10,17,26,35,36,39,12)]
colnames(bp_fus_sh)<-header
bp_fis_sh<-bp_fis[,c(2,37,38,10,17,26,35,36,39,12)]
colnames(bp_fis_sh)<-header

#tips
bp_fus_F_sh<-bp_fus_FMR[,c(4,35,36,38,40,24,33,34,37,2)]
colnames(bp_fus_F_sh)<-header
bp_fis_F_sh<-bp_fis_FMR[,c(4,35,36,38,40,24,33,34,37,2)]
colnames(bp_fis_F_sh)<-header

#duplicate for male or Female in tips 
bp_fus_M_sh<-bp_fus_FMR[,c(4,35,36,39,40,24,33,34,37,3)]
colnames(bp_fus_M_sh)<-header
bp_fis_M_sh<-bp_fis_FMR[,c(4,35,36,39,40,24,33,34,37,3)]
colnames(bp_fis_M_sh)<-header

#Male tips need to switch polarity of fis and fus

bp_fis_M_sh$BP<-str_replace_all(bp_fis_M_sh$BP, "Fis", "Fus")
bp_fus_M_sh$BP<-str_replace_all(bp_fus_M_sh$BP, "Fus", "Fis")


breakpoints_tmp<-rbind(bp_fis_sh,bp_fus_sh, bp_fis_F_sh, bp_fus_F_sh, bp_fis_M_sh, bp_fus_M_sh)

View(breakpoints_tmp)

breakpoints<-breakpoints_tmp %>% 
  group_by(across(all_of(c("BP","NODE","SPECIES")))) %>%
  filter(n() == 2) %>%
  mutate(START=as.numeric(START),
         STOP=as.numeric(STOP))%>%
  ungroup()

#View(breakpoints)

#Filter the BPs 

#Overlapp intervals and give them a ID based on overlapping ranges on same chromsomes


gr <- GRanges(
  seqnames = breakpoints$CHROM,
  ranges   = IRanges(start = breakpoints$START, end = breakpoints$STOP)
)


gr_red <- reduce(gr, with.revmap = TRUE)

overlap_id <- integer(length(gr))

for (i in seq_along(gr_red)) {
  overlap_id[ gr_red$revmap[[i]] ] <- i
}

breakpoints$overlap_id <- overlap_id

#View(breakpoints)

#Connect both breakpoints with a common name.
breakpoints_stdN<-data.frame()
for(i in 1:length(unique(breakpoints$SPECIES))){
  for(j in 1:length(unique(breakpoints$BP))){
    for(l in 1:length(unique(breakpoints$NODE))){
      d<-breakpoints[breakpoints$SPECIES %in% unique(breakpoints$SPECIES)[i] & 
                       breakpoints$BP %in% unique(breakpoints$BP)[j] &
                       breakpoints$NODE %in% unique(breakpoints$NODE)[l] ,]
      
      sh<-unique(str_remove_all(d$BP, "\\d*" ))
      id<-c(as.numeric(d$overlap_id))
      d$stdBP<-paste0(sh,"_", min(id), "-", max(id) )
      
      breakpoints_stdN<-rbind(breakpoints_stdN, d)
    }
  }
}

#View(breakpoints_stdN)


```

#EBR placement on the phylogeny
```{r}


breakpoints_stdN$combID<-str_remove_all(str_remove_all(breakpoints_stdN$stdBP, "Fis_"), "Fus_")


#View(breakpoints_stdN)


#Polarise polarise N5; Sinapis Swe
#These are from N5 which are not included in agora, hence I polarise them based on N3 
#13-13 Fis sweF
#4-23 Fus sweF
#38-12 Fus sweF
#22-22 Fis sweM
#26-6 Fis sweF
#26-6 Fus sweF

sweFN5<-c("59-69", "83-83", "142-167", "100-121") #Should be removed in sinapissweM N5
sweMN5<-"79-79"  #Should be removed in sinapissweF N5

breakpoints_stdN_tmp<-breakpoints_stdN[!(breakpoints_stdN$SPECIES == "LsinapisSweM" & 
                                         breakpoints_stdN$NODE == "N5" &
                                         breakpoints_stdN$combID %in% sweFN5) & 
                                         !(breakpoints_stdN$SPECIES == "LsinapisSweF" & 
                                         breakpoints_stdN$NODE == "N5" &
                                         breakpoints_stdN$combID %in% sweMN5) , ]




#Filter based on phylogeny
BP_phylo<-breakpoints_stdN_tmp %>% 
        group_by(stdBP) %>%
        mutate(highestNODE = as.numeric(str_remove_all(NODE, "N")),
               split = str_remove_all(BP, "\\d*"),
               CHROM = str_replace_all(CHROM, "Z", "48")) %>%
        ungroup() %>%
        group_by(combID, SPECIES) %>%
        filter(highestNODE == max(highestNODE)) %>%
        ungroup()  %>%
        group_by(combID, SPECIES) %>%
  ungroup() 

View(BP_phylo)

BP_phylo$id1<-str_split(BP_phylo$combID, "-", simplify = T)[,1]
BP_phylo$id2<-str_split(BP_phylo$combID, "-", simplify = T)[,2]



df_long <- BP_phylo %>%
  pivot_longer(cols = c(id1, id2),
    names_to = "id_pos",
    values_to = "id")




id_class <- df_long %>%
  group_by(SPECIES, id) %>%
  summarise(n_combID = n_distinct(combID),
    n_NODE   = n_distinct(NODE),
      reuse_decendent_tmp = case_when(
      n_combID >= 2 & n_NODE > 1 ~ "reuse_decendent",
      TRUE ~ NA_character_  ),
    translocation_tmp = case_when(
      n_combID >= 2 & n_NODE == 1 ~ "translocation",
      TRUE ~ NA_character_  ),
    .groups = "drop")




id_class$overlap_id<-as.numeric(id_class$id)

df<-left_join(BP_phylo, id_class, by = c("SPECIES", "overlap_id") )



df2<-df  %>%
        group_by(combID, SPECIES) %>%
          mutate(reuse_decendent = case_when(
            any(reuse_decendent_tmp == "reuse_decendent") ~ "reuse_decendent",
            TRUE ~ NA),
            translocation = case_when(
            any(translocation_tmp == "translocation") ~ "translocation",
            TRUE ~ NA)
          ) %>%
        ungroup() %>%
        group_by(combID, SPECIES, NODE) %>%
        mutate(reciprocal_translocation = case_when(
            any(length(combID) == 4 ) ~ "reciprocal_translocation",
            TRUE ~ NA)
          ) %>%
  ungroup()
  


df3<-df2[,c(1,2,3,4,5,12,7,8,13,15,23,24,25,10)] 


colnames(df3)[10]<-"EBR_TYPE"
colnames(df3)[11]<-"reuse_decendent"
colnames(df3)[12]<-"translocation"




df3$EBR_TYPE<-case_when(df3$EBR_TYPE == "Fis" & !df3$NODE %in% c("N0", "N2")  ~ "Fission", 
                      df3$EBR_TYPE == "Fus"  & !df3$NODE %in% c("N0", "N2") ~ "Fusion",
            TRUE ~ "Unpolarisable")




#Remove EBRs on sex chromosomes that does not make sense and are likely due to misassembly, these are removed as a conserative mesure
df4 <- df3[!(df3$SPECIES %in% c("LsinapisSweF","LsinapisSweM", "LsinapisCatM", "LsinapisCatF", "LjuvernicaSweF","LjuvernicaSweM", "LsinapisAstM" ) & df3$CHROM == "Chr_3"), ]





#Branch specific EBRs

df5<-df4 %>%
  mutate(BRANCH = case_when(
             (SPECIES %in% c("LjuvernicaSweF","LjuvernicaSweM")  & NODE == "N0") ~ "AncAll-AncLjuvernica",
             (!SPECIES %in% c("LjuvernicaSweF","LjuvernicaSweM")  & NODE == "N0") ~ "AncAll-AncLsinapis/Lreali",
             (SPECIES %in% c("LjuvernicaSweF")  & NODE == "N2") ~ "AncLjuvernica-LjuvernicaSweF",
             (SPECIES %in% c("LjuvernicaSweM")  & NODE == "N2") ~ "AncLjuvernica-LjuvernicaSweM",
             (SPECIES %in% c("LrealiCatF","LrealiCatM")  & NODE == "N1") ~ "AncLsinapis/Lreali-LrealiCat",
             (!SPECIES %in% c("LrealiCatF","LrealiCatM")  & NODE == "N1") ~ "AncLsinapis/Lreali-AncLsinapis",
             (SPECIES %in% c("LsinapisSweF","LsinapisSweM")  & NODE == "N3") ~ "AncLsinapis-AncLsinapisSwe",
             (SPECIES %in% c("LsinapisSweF")  & NODE == "N5") ~ "AncLsinapisSwe-LsinapisSweF",
             (SPECIES %in% c("LsinapisSweM")  & NODE == "N5") ~ "AncLsinapisSwe-LsinapisSweM",
             (!SPECIES %in% c("LsinapisSweF","LsinapisSweM")  & NODE == "N3") ~ "AncLsinapis-AncLsinapisIberian",
             (SPECIES %in% c("LsinapisCatF","LsinapisCatM")  & NODE == "N6") ~ "AncLsinapisIberian-LsinapisCat",
             (!SPECIES %in% c("LsinapisCatF","LsinapisCatM")  & NODE == "N6") ~ "AncLsinapisIberian-LsinapisAst",
            TRUE ~ NA),
  ) %>%
  ungroup()


View(df5)

#Write the data frame
write_delim(df5, file = "leptidea.breakpoints.BranchEBRs.withOrgChroms.tsv", delim = "\t")

```
 


#Summarise EBRs per branch

```{r}

counts <- df5 %>%
  group_by(BRANCH, EBR_TYPE) %>%
  summarise(n = n()/2, 
            nSPECIES = n_distinct(SPECIES),
            .groups = "drop"
            )

View(counts)


counts_S <- df5 %>%
  group_by(SPECIES, EBR_TYPE, BRANCH) %>%
  summarise(n = n()/2, 
            nSPECIES = n_distinct(SPECIES),
            .groups = "drop"
            )

View(counts_S)


#Count translocations
counts_trans <- df5 %>%
  group_by(BRANCH, translocation) %>%
  summarise(n = n()/2, 
            nSPECIES = n_distinct(SPECIES),
            .groups = "drop"
            )

View(counts_trans)

#Count reciprocal translocations
counts_recitrans <- df5 %>%
  group_by(BRANCH, reciprocal_translocation) %>%
  summarise(n = n()/2, 
            nSPECIES = n_distinct(SPECIES),
            .groups = "drop"
            )



#Largest EBR region
max(df5[df5$EBR_TYPE %in% c("Fission", "Fussion"),]$STOP-df5[df5$EBR_TYPE %in% c("Fission", "Fussion"),]$START)

#Smallest EBR region
min(df5[df5$EBR_TYPE %in% c("Fission", "Fussion"),]$STOP-df5[df5$EBR_TYPE %in% c("Fission", "Fussion"),]$START)

```


#Busco phylogeny with rates of EBRs

```{r}

#Plot the phylogeny
p<-ggtree(tree)+ 
  geom_tiplab(hjust = -.01)+
  geom_nodelab(hjust = -1, size=3)+
  theme_tree2() + 
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) + 
  theme_tree()+
  geom_text2(aes(label = round(branch.length, 4)),
             hjust = 1.2,
             vjust = -0.3,
             size = 3)

p

#Extract the names of parent and child nodes to have unique branch names to transfer EBR data on
edge_df <- data.frame(
  edge_id = 1:nrow(tree$edge),
  parent = tree$edge[,1],
  child  = tree$edge[,2],
  branch_length = tree$edge.length
)

#This order is set manually based on the phylogeny. 
edge_df$BRANCH<-c("AncAll-AncLsinapis/Lreali", 
                      "AncLsinapis/Lreali-LrealiCat",
                      "AncLreali/LrealiCatF",
                      "AncLreali/LrealiCatM",
                      "AncLsinapis/Lreali-AncLsinapis",
                      "AncLsinapis-AncLsinapisIberian",
                      "AncLsinapisIberian-LsinapisCat",
                      "LsinapisCat-LsinapisCatM",
                      "LsinapisCat-LsinapisCatF",
                      "AncLsinapisIberian-LsinapisAst",
                      "AncLsinapis-AncLsinapisSwe",
                      "AncLsinapisSwe-LsinapisSweF",
                      "AncLsinapisSwe-LsinapisSweM",
                      "AncAll-AncLjuvernica",
                      "AncLjuvernica-LjuvernicaSweM",
                      "AncLjuvernica-LjuvernicaSweF")


#Fissions
count_fis<-counts[counts$EBR_TYPE == "Fission",]
merged_df_fis <- left_join(edge_df, count_fis, by = "BRANCH")

merged_df_fis$rate<-(merged_df_fis$n/merged_df_fis$nSPECIES)*merged_df_fis$branch_length

merged_df_fis[c(1,14,15,16),]$rate<-"NA" #Set the branches that are unpolarisable to NA; set manually
merged_df_fis[c(3,4,8,9,11),]$rate<-0 # Assign branches without any events to 0 instead of NA; set manually

#Attach fission rates to the tree file
rate_df_fis <- data.frame(
  node = tree$edge[,2],   # child nodes
  fission_rate = as.numeric(merged_df_fis$rate),
  events = as.numeric(merged_df_fis$n/merged_df_fis$nSPECIES)
)


#Fusions
count_fus<-counts[counts$EBR_TYPE == "Fusion",]
merged_df_fus <- left_join(edge_df, count_fus, by = "BRANCH")

merged_df_fus$rate<-(merged_df_fus$n/merged_df_fus$nSPECIES)*merged_df_fus$branch_length

merged_df_fus[c(1,14,15,16),]$rate<-"NA" #Set the branches that are unpolarisable to NA; set manually
merged_df_fus[c(3,4,8,9,10,13),]$rate<-0 # Assign branches without any events to 0 instead of NA; set manually

#Attach fusion rates to the tree file
rate_df_fus <- data.frame(
  node = tree$edge[,2],   # child nodes
  fusion_rate = as.numeric(merged_df_fus$rate),
  events = as.numeric(merged_df_fus$n/merged_df_fus$nSPECIES)
)


#Translocations
count_trans<-counts_trans[counts_trans$translocation == "translocation",]
merged_df_trans <- left_join(edge_df, count_trans, by = "BRANCH")

merged_df_trans$rate<-(merged_df_trans$n/merged_df_trans$nSPECIES)*merged_df_trans$branch_length

count_rtrans<-counts_recitrans[counts_recitrans$reciprocal_translocation == "reciprocal_translocation",]
merged_df_rtrans <- left_join(edge_df, count_rtrans, by = "BRANCH")

merged_df_rtrans$rate<-(merged_df_rtrans$n/merged_df_rtrans$nSPECIES)*merged_df_rtrans$branch_length

merged_df_trans[c(1,14,15,16),]$rate<-"NA" #Set the branches that are unpolarisable to NA; set manually
merged_df_trans[c(3,4,7,8,9,10,11,13),]$rate<-0 # Assign branches without any events to 0 instead of NA; set manually

merged_df_trans[c(12),]$rate<-merged_df_rtrans[c(12),]$rate # add reciprocal translocations
merged_df_trans[c(12),]$n<-merged_df_rtrans[c(12),]$n # add reciprocal translocations
merged_df_trans[c(12),]$nSPECIES<-merged_df_rtrans[c(12),]$nSPECIES # add reciprocal translocations

#Attach translocation rates to tree
rate_df_trans <- data.frame(
  node = tree$edge[,2],   # child nodes
  trans_rate = as.numeric(merged_df_trans$rate),
  events = as.numeric(merged_df_trans$n/merged_df_trans$nSPECIES)
)

#For combined header
rate_scale <- scale_color_gradientn(
  name = "EB rate",
    colours = c("grey70", "yellow", "orange","red", "red4"),
    values = scales::rescale(c(0, 0.0001, 0.03, 0.075, 0.15)),
  limits = c(0, 0.15),
  oob = scales::squish,
  na.value = "black"
)

legend <- cowplot::get_legend(
  ggtree(tree) %<+% rate_df_fis +
    geom_tree(aes(color = fission_rate), size = 1) +
    rate_scale +
    theme(legend.position = "right")
)

pfis <- ggtree(tree) %<+% rate_df_fis +
  geom_tree(aes(color = fission_rate), size = 2, lineend = "square") +
  rate_scale +
  theme_tree() +
  geom_tiplab(hjust = -0.1, size = 2) +
  geom_text2(aes(label = events), hjust = 1.8, vjust= -0.5, size=2.1)+ #Events on branches
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.4)))+ 
  ggtitle("     Fissions") +
  theme(legend.position = "none",
        title = element_text(size=8))



pfus<-ggtree(tree) %<+% rate_df_fus +
  geom_tree(aes(color = fusion_rate), size = 2, lineend = "square") +
  rate_scale +
  theme_tree2() +
  geom_tiplab(hjust = -0.1, size = 3) +
  geom_text2(aes(label = events), hjust = 1.5, vjust= -0.5, size=2.5)+ #Events on branches
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.4)))+ 
  ggtitle("     Fusions") +
  theme(title = element_text(size=8),
  legend.position = c(0.98, 0.02),   # bottom right (inside plot)
  legend.justification = c(1, 0),
  legend.key.height = unit(0.4, "cm"),
  legend.key.width  = unit(0.4, "cm"),
  legend.ticks = element_line(colour="black")
)


ptrans<-ggtree(tree) %<+% rate_df_trans +
  geom_tree(aes(color = trans_rate), size = 2, lineend = "square") +
  rate_scale +
  theme_tree() +
   geom_text2(aes(label = events), hjust = 1.8, vjust= -0.5, size=2.1)+ #Events on branches
  geom_tiplab(hjust = -0.1, size = 2) +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.4)))+ 
  ggtitle("     Translocations") +
  theme(legend.position = "none",
        title = element_text(size=8))



trees_vertical <- cowplot::plot_grid(
  pfis,
  ptrans,
  ncol = 1,
  align = "v",
  axis = "lr",
  labels = c("B","C")
)




final_plot <- cowplot::plot_grid(
  pfus,
  trees_vertical,
  ncol = 2,
  labels= c("A", ""),
  rel_widths = c(0.65, 0.35)
)

final_plot

#ggsave("../Busco_phylo/Fig1.pdf",final_plot,  height = 120, width= 180, units = "mm", dpi = 600)
```
